---
title: "Effects of wheather phenomena in the USA on population health and economincs"
author: "Daniel Jungen"
date: "29 Oktober 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache =  TRUE)
```

## Effects of weather phenomena on health and economics

In this report we address the questions:

- Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

- Across the United States, which types of events have the greatest economic consequences?

The analysis is based on data of the NOAA Storm Database. The datafile is assumed to be downloaded to the working directory of R and namend "data.csv".

It can be found on <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2>.

# Summary

```{r echo =TRUE}
library(dplyr)

storm_data<-read.csv("data.csv")

pop_health<-data.frame(cbind(data.frame(storm_data$EVTYPE), data.frame(storm_data$FATALITIES), data.frame(storm_data$INJURIES)))

pop_health_fat<-aggregate(pop_health$storm_data.FATALITIES, by=list(evtype=pop_health$storm_data.EVTYPE), FUN=sum )
pop_health_inj<-aggregate(pop_health$storm_data.INJURIES, by=list(evtype=pop_health$storm_data.EVTYPE), FUN=sum )

names(pop_health_fat)<-c("EVTYPE", "FATALITIES")
names(pop_health_inj)<-c("EVTYPE", "INJURIES")

pop_health_agg<-merge(pop_health_fat, pop_health_inj, by= "EVTYPE")

pop_health_agg<-mutate(pop_health_agg, sum_casualties = pop_health_agg$FATALITIES+pop_health_agg$INJURIES)

temp<-pop_health_agg[order(-pop_health_agg$FATALITIES),]

```

The top ten events on population health in termes of fatalities with the corresponding numbers are 

```{r echo = TRUE}
temp[1:10,1:2]
```

The same with respect to injuries returns:

```{r echo = TRUE}

temp<-pop_health_agg[order(-pop_health_agg$INJURIES),]
temp[1:10,c(1,3)]
```

And taking the sum of casualties in fatalities and injuries we get

```{r echo = TRUE}
temp<-pop_health_agg[order(-pop_health_agg$sum_casualties),]
temp[1:10,c(1,4)]
```

#Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r echo = TRUE}
library(dplyr)

#storm_data<-read.csv("data.csv")

pop_health<-data.frame(cbind(data.frame(storm_data$EVTYPE), data.frame(storm_data$FATALITIES), data.frame(storm_data$INJURIES)))

pop_health_fat<-aggregate(pop_health$storm_data.FATALITIES, by=list(evtype=pop_health$storm_data.EVTYPE), FUN=sum )
pop_health_inj<-aggregate(pop_health$storm_data.INJURIES, by=list(evtype=pop_health$storm_data.EVTYPE), FUN=sum )

names(pop_health_fat)<-c("EVTYPE", "FATALITIES")
names(pop_health_inj)<-c("EVTYPE", "INJURIES")

pop_health_agg<-merge(pop_health_fat, pop_health_inj, by= "EVTYPE")

pop_health_agg<-mutate(pop_health_agg, sum_casualties = pop_health_agg$FATALITIES+pop_health_agg$INJURIES)

top_fatalities<-data.frame(pop_health_agg[order(-pop_health_agg$FATALITIES),])

top_fatalities[1:10,c(1,2)]

top_injuries<-data.frame(pop_health_agg[order(-pop_health_agg$INJURIES),])

top_injuries<-top_injuries[1:10,c(1,3)]

top_casualties<-data.frame(pop_health_agg[order(-pop_health_agg$sum_casualties),])

top_casualties<-top_casualties[1:10,c(1,4)]

```
# Across the United States, which types of events have the greatest economic consequences?

Just like in the section above we now determine the damage that was caused by different weather phenomena.

```{r}
damages<-data.frame(cbind(data.frame(storm_data$EVTYPE), data.frame(storm_data$PROPDMG), data.frame(storm_data$CROPDMG)))

damages_prop<-aggregate(damages$storm_data.PROPDMG, by=list(evtype=damages$storm_data.EVTYPE), FUN=sum )
damages_crop<-aggregate(damages$storm_data.CROPDMG, by=list(evtype=damages$storm_data.EVTYPE), FUN=sum )

names(damages_prop)<-c("EVTYPE", "PROPDMG")
names(damages_crop)<-c("EVTYPE", "CROPDMG")

damages_agg<-merge(damages_prop, damages_crop, by= "EVTYPE")

damages_agg<-mutate(damages_agg, sum_damage = damages_agg$PROPDMG + damages_agg$CROPDMG)

top_prop<-data.frame(damages_agg[order(-damages_agg$PROPDMG),])

top_prop[1:10,c(1,2)]

top_crop<-data.frame(damages_agg[order(-damages_agg$CROPDMG),])

top_crop[1:10,c(1,3)]

top_sum_damage<-data.frame(damages_agg[order(-damages_agg$sum_damage),])

top_sum_damage<-top_sum_damage[1:10,c(1,4)]
```

